TSELM: Target Speaker Extraction using Discrete Tokens and Language Models

### 2024-09-12 src/paper.tex
update

\begin{IEEEkeywords}
target speaker extraction, speech separation, language models, audio discretization, self-supervised models
\end{IEEEkeywords}
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.77\textwidth]{assets/model.pdf}
    \caption{Overview of TSELM.}
    \label{model}
    \vspace{-12pt}
    \end{figure*}

    \begin{figure}
        \centering
        \includegraphics[width=0.25\textwidth]{assets/cross_attention.pdf}
        \caption{Details of Cross-Attention mechanism.}
        \label{cross_attention}
        \vspace{-17pt}
        \end{figure}

@@

There are only few works about discretization in Target Speech Extraction (TSE). SkiM-UniCATS \cite{gen_tse} is among the first to use discrete tokens in TSE, leveraging SSL models like HuBERT and vq-wav2vec \cite{vq_wav2vec}. However, it overlooks the WavLM model, which excels in speech separation. Also, it mainly focuses on single-layer outputs for discretization. Moreover, its evaluation is limited to speech quality, neglecting intelligibility and speaker similarity, both crucial for TSE. We try to address these potential issues by proposing TSELM, which uses multiple WavLM layers for discretization and adds metrics for intelligibility and speaker similarity. Our demos are available at 
\footnote{\href{https://beilong-tang.github.io/TSELM.demo/}{https://beilong-tang.github.io/TSELM.demo/}}.

\section{Method}
TSELM has three stages: encoding, modeling and decoding. In the encoding stage, both reference and mixed speech are tokenized using WavLM and Kmeans. The reference speech is passed directly to the encoder, while for mixed speech, we concatenate the reference speech to both sides of the mixture before passing it through the WavLM model. After tokenization, we retain only the tokens corresponding to the mixed speech.
In the modeling stage, an attention embedding mechanism is employed to combine the embeddings from all layers. A cross-attention mechanism, similar to that used in \cite{usef_tes}, is applied to inject speaker-specific information. An encoder-only language model, followed by a linear classifier, is then used to generate the reconstructed tokens.
In the decoding stage, we leverage the pretrained scalable HiFi-GAN from \cite{unit_hifi} to reconstruct the audio from the discrete tokens. Unlike SELM \cite{selm}, where a conformer detokenizer is trained to reconstruct WavLM embeddings before passing them through HiFi-GAN, the scalable HiFi-GAN in \cite{unit_hifi} uses a dropout mechanism to directly reconstruct audio from multiple layers of tokens, eliminating the need for a conformer detokenizer and the complexity of training a separate HiFi-GAN for each layer. Both the encoder and decoder are kept frozen during training, and an overview of the model is shown in Fig.\ref{model}.
Extensive experiments demonstrate that our method achieves excellent speech quality and comparable intelligibility results.

@@

We use the pretrained self-supervised learning (SSL) model WavLM Large \cite{wavlm} to encode speech into continuous representations. Specifically, we extract the outputs from six hidden layers: 1, 3, 7, 12, 18, and 23. Given a speech signal \(s \in \mathbb{R}^{T'}\), the output of WavLM is a tensor \(\bm{r}\) with shape \(n \times T \times E\), where \(n\) is the number of output layers (6 in this case), \(T\) is the time dimension, and \(E\) represents the embedding dimension. For tokenization, we apply a separate Kmeans model to each output layer, with each model using the same number of clusters, denoted by \(K\). After tokenization, the continuous embedding \(\bm{r}\) is transformed into a discrete tensor \(\bm{d}\) with shape \(n \times T\), where each value \(\bm{d}_{i} \in (0, K-1) \). In all our experiments, we set \(K = 1000\). For both reference and mixed speech, the same Kmeans model and the same layer combination from WavLM Large are used. The encoder remains frozen during training.
The encoding strategy for mixed speech is crucial to the performance of the model. Given a reference speech \(s_r \in \mathbb{R}^{T^r}\) and a mixed speech \(s_m \in \mathbb{R}^{T'}\), we follow the previously described procedure to encode the reference speech into a tensor \(\bm{d_r}\) of shape \(n \times T_r\). However, for the mixed speech, instead of applying the encoding directly, we first concatenate it with the reference speech, creating a signal \(s' = [s_r, s_m, s_r] \in \mathbb{R}^{(T^r + T' + T^r)}\). This concatenated signal is then input into the encoder, producing an output tensor \(\bm{d'}\) with shape \(n \times (T_r + T + T_r)\), where \(T\) represents the length for the mixed speech embedding. The tensor \(\bm{d'}\) contains discrete tokens for the two segments of reference speech and the mixed speech. We extract the portion \(\bm{d}\) corresponding to the mixed speech, resulting in an output tensor of shape \(n \times T\).

@@

  \caption{Results on Libri2Mix (clean) and WSJ0-2mix: In the "Category" column, "G" refers to generative models, while "D" refers to discriminative models. The "Type" column categorizes methods as "D" (discrete), "H" (hybrid), or "C" (continuous).
  For discrete methods, speaker similarity is compared against the discretized target speech (denoted as Target-discrete-WavLM) instead of the original target speech. It is denoted as "\_d".}

@@

@@

  SkiM-UniCATS\cite{gen_tse}                   & G                                             & D                                                                                  & -        & -      & -       & -            & -     &          & 3.62        & 4.10       & 3.37       & -            & -             \\ \hline

@@

We utilize the output from hidden layers 1, 3, 7, 12, 18, 23 from WavLM Large and Kmeans model with 
\(K=1000\).
We use 1024 as the embedding dimension.  The cross-attention module consists of four transformer encoders, each with 16 attention heads and an MLP with a hidden dimension of 1024. Layer normalization is applied after the cross-attention module.
The LM of small version TSELM-S uses 
embedding dimension \(d\) = 256, absolute sinusoidal positional embedding, conformer encoders as backbone of LM. The conformer encoder consists of 6 layers with a kernel size of 31, each with 4 attention heads, and an MLP with a hidden dimension of 2048. The medium version TSELM-M uses \(d\) = 512 with 8 layers and 8 heads and the large version TSELM-L uses 
\(d\) = 768 with 12 layers and 16 heads. We use AdamW as 
our optimizer for all the experiments. The learning rate 
is \(5 \times 10^{-4}\) for TSELM-S and \(5 \times 10^{-5}\) for TSELM-M and TSELM-L. We train the model using 8 GPUs with 32 GB of RAM, each with batch size 16 for a total of 40,000 steps.

@@

\subsection{Baseline models}
Baseline models are presented in Table \ref{main_exp}. 
TSELM is compared with Spex+ \cite{spex_plus}, a discriminative separation model trained on Libri2Mix \cite{librimix}. 
We also compare our model with SkiM-UniCATS(vq-wav2vec) from \cite{gen_tse}, which uses discrete tokens from vq-wav2vec to conduct TSE. 
Mixture refers to the unprocessed mixed speech. 
Target-Discrete refers to the discretized target speech using our encoder and decoder. 
It serves as the upper bond for our model performance.  Besides TSELM, we conduct three main experiments, named
 Continuous-WavLM-L6, TSELM-L-Hybrid and TSELM-S-NoCat using the same training data. 
For Continuous-WavLM-6, we directly
pass the embeddings from the 6th hidden layer output of WavLM Large to the cross- attention  
and LM without discretization. 
The concatenation strategy is still applied to the mixed speech. Mean Square Error (MSE) loss is applied between the output embeddings and the clean 
embeddings. HiFi-GAN in \cite{knn_vc} is used for audio reconstruction. For 
TSELM-L-Hybrid, inspired by MaskSR \cite{mask_sr}, we discretize the reference 
speech while utilizing the continuous embeddings from the mixed speech. We call 
it hybrid because the mixture speech retains continuous features. In TSELM-S-NoCat, we utilize the TSELM-S model architecture but abandon the mixture concatenation strategy in the encoding step. The code and pretrained models are available at \footnote{\href{https://github.com/Beilong-Tang/TSELM}{https://github.com/Beilong-Tang/TSELM}}.

\section{Results and Discussions}
Table \ref{main_exp} presents the performance of different systems evaluated on the Libri2Mix test and WSJ0-2mix testset. 
DNSMOS is computed over the 
output since this metric is reference-free. dWER is calculated with respect to the clean speech. 
For continuous methods, speaker similarity is directly computed against the clean speech. 
However, since discretization inherently results in some loss of speaker information—Target-Discrete shows a speaker similarity score of 0.653 on Libri2Mix —we follow \cite{dasb} to assess the speaker similarity of the output from the discrete methods against the target audio produced by discretizing the clean speech. 
This target audio output from Target-Discrete serves as the upper bond for our model performance.
The observed speaker information loss is likely due to the tokenization process, which inherently reduces speaker fidelity. Future work should aim to enhance tokenization methods for SSL models to mitigate this loss. Since the primary goal of this research is to explore target speaker extraction using discretized information rather than to develop improved tokenization methods, we argue that comparing the outputs of our model with discretized speech is a reasonable approach, as it represents an upper bound on performance.

We observe that TSELM-L outperforms Spex+ in terms of DNSMOS scores, indicating better speech quality, but performs slightly worse in dWER, suggesting lower speech intelligibility. One potential reason for this could be the discretization process applied to the mixed speech. Our Kmeans algorithm is trained on clean speech rather than mixed speech, which is advantageous for speech enhancement as it likely aids in denoising. However, when applied to mixed speech, this discretization might lead to the model focusing on the wrong speaker, as it may only retain dominant speaker information, causing a reduction in intelligibility.
This hypothesis is supported by the results of TSELM-L-Hybrid, where continuous embeddings from the mixture speech were used without discretization, achieving dWER scores better than Spex+ on WSJ0-2mix. 
% Another contributing factor could be the limitations of our current language model (LM). The encoder-only LM achieves around 55\% accuracy for token prediction on Libri2Mix testset, and we believe using more advanced models, such as auto-regressive or masking-based LMs, could lead to improved performance in future iterations.

@@

In Table \ref{hubert_wavlm_6}, we compare the performance of HuBERT and WavLM as SSL models and examine the effects of using either one or multiple layers for discretization. Our findings indicate that using HuBERT as the SSL model results in better DNSMOS scores but much worse dWER compared to our WavLM baseline. The improved DNSMOS scores likely stem from the vocoder performance, yet they do not adequately reflect speech intelligibility, which is crucial for speech separation tasks.
The poorer dWER scores observed with this HuBERT model may be attributed to its training on clean speech, which might not equip it to capture the complexity and richness of mixed speech. Moreover, our results from WavLM-L6 suggest that when performing speech separation, discretizing across multiple layers provides better results than relying on a single layer.  This might be because that using multiple layer outputs can better tolerate errors compared to using just one layer output.

Finally, we observe a performance gap between discrete methods and continuous methods, as demonstrated by Continuous-WavLM-L6.
Continuous-WavLM-L6 has the best performance in terms of DNSMOS and dWER among all the experiments, and it only uses the 6 layer output of WavLM. 
The gap in performance may be attributed to the information loss inherent in the discretization process. 
We hope our future research will bridge this gap.

\section{Conclusion}
In this work, we introduce a novel way using discrete tokens and language models for target speaker extraction.
Our method leverages multiple hidden layers of WavLM and Kmeans tokenizers for encoding, employs cross-attention and a language model for separation, and utilizes a scalable HiFi-GAN for audio reconstruction.
Experiments have 
shown that our model can achieve excellent performance in terms of speech quality, and comparable performance in terms of speech intelligibility. However, we observe a gap between discrete methods and continuous methods especially in speech intelligibility and speaker similarity. We will focus on shrinking 
this gap.

>>>>

\begin{IEEEkeywords}
target speaker extraction, speech separation, language models, audio discretization, self-supervised models
\end{IEEEkeywords}
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.77\textwidth]{assets/model.pdf}
    \caption{Overview of our proposed target speaker extraction framework with discrete tokens and language models.}
    \label{model}
    \vspace{-12pt}
    \end{figure*}

    \begin{figure}
        \centering
        \includegraphics[width=0.25\textwidth]{assets/cross_attention.pdf}
        \caption{Details of the Cross-Attention mechanism in modeling.}
        \label{cross_attention}
        \vspace{-17pt}
        \end{figure}

@@

There are only few works about discretization in Target Speech Extraction (TSE). SkiM-UniCATS \cite{gen_tse} is among the first to use discrete tokens in TSE, leveraging SSL models like HuBERT and vq-wav2vec \cite{vq_wav2vec}. However, it overlooks the WavLM model, which excels in speech separation \cite{wavlm}. Also, it mainly focuses on single-layer outputs for discretization. Moreover, its evaluation is limited to speech quality, neglecting intelligibility and speaker similarity, both crucial for TSE. We try to address these potential issues by proposing TSELM, which uses multiple WavLM layers for discretization and adds metrics for intelligibility and speaker similarity. Our demos are available at 
\footnote{\href{https://beilong-tang.github.io/TSELM.demo/}{https://beilong-tang.github.io/TSELM.demo/}}.

\section{Method}
TSELM has three stages: encoding, modeling and decoding. In the encoding stage, both reference and mixed speech are tokenized using WavLM and Kmeans. The reference speech is passed directly to the encoder, while for the mixed speech, we concatenate the reference speech to both sides of the mixture before passing it through the WavLM model. After tokenization, we retain only the tokens corresponding to the mixed speech.
In the modeling stage, an attention embedding mechanism is employed to combine the embeddings from all layers. A cross-attention mechanism, similar to that used in \cite{usef_tes}, is applied to inject speaker-specific information. An encoder-only language model, followed by a linear classifier, is then used to generate the reconstructed tokens.
In the decoding stage, we leverage the pretrained scalable HiFi-GAN in \cite{unit_hifi} to reconstruct the audio from the discrete tokens. Unlike SELM \cite{selm}, where a conformer detokenizer is trained to reconstruct WavLM embeddings before passing them through HiFi-GAN, the scalable HiFi-GAN in \cite{unit_hifi} uses a dropout mechanism to directly reconstruct audio from multiple layers of tokens, eliminating the need for a conformer detokenizer and the complexity of training a separate HiFi-GAN for each layer. Both the encoder and decoder are kept frozen during training. An overview of TSELM is shown in Fig.\ref{model}.
Extensive experiments demonstrate that our method achieves excellent speech quality and comparable intelligibility results.

@@

We use the pretrained SSL model WavLM Large \cite{wavlm} to encode speech into continuous representations. Specifically, we extract the outputs from six hidden layers: 1, 3, 7, 12, 18, and 23. Given a speech signal \(s \in \mathbb{R}^{T'}\), the output of WavLM is a tensor \(\bm{r}\) with shape \(n \times T \times E\), where \(n\) is the number of output layers (6 in this case), \(T\) is the time dimension, and \(E\) represents the embedding dimension. For tokenization, we apply a separate Kmeans model to each output layer, with each model using the same number of clusters, denoted by \(K\). After tokenization, the continuous embedding \(\bm{r}\) is transformed into a discrete tensor \(\bm{d}\) with shape \(n \times T\), where each value \(\bm{d}_{i} \in (0, K-1) \). In all our experiments, we set \(K = 1000\). For both reference and mixed speech, the same Kmeans model and the same layer combination from WavLM Large are used. The encoder remains frozen during training.

The encoding strategy for mixed speech is crucial to the performance of the model. Given a reference speech \(s_r \in \mathbb{R}^{T^r}\) and a mixed speech \(s_m \in \mathbb{R}^{T'}\), we follow the previously described procedure to encode the reference speech into a tensor \(\bm{d_r}\) of shape \(n \times T_r\). However, for the mixed speech, instead of applying the encoding directly, we first concatenate it with the reference speech, creating a signal \(s' = [s_r, s_m, s_r] \in \mathbb{R}^{(T^r + T' + T^r)}\). This concatenated signal is then input into the encoder, producing an output tensor \(\bm{d'}\) with shape \(n \times (T_r + T + T_r)\), where \(T\) represents the length for the mixed speech embedding. The tensor \(\bm{d'}\) contains discrete tokens for the two segments of reference speech and the mixed speech. We extract the portion \(\bm{d}\) corresponding to the mixed speech, resulting in an output tensor of shape \(n \times T\).

@@

  \caption{Results on Libri2Mix clean and WSJ0-2mix: In the "Category" column, "G" refers to generative models, while "D" refers to discriminative models. The "Type" column categorizes methods as "D" (discrete), "H" (hybrid), or "C" (continuous).
  For discrete methods, speaker similarity is compared against the discretized target speech (denoted as Target-Discrete) instead of the original target speech. It is denoted as "\_d".}

@@

    \setlength{\tabcolsep}{4.9pt}

@@

  SkiM-UniCATS(vq-wav2vec)\cite{gen_tse}                   & G                                             & D                                                                                  & -        & -      & -       & -            & -     &          & 3.62        & 4.10       & 3.37       & -            & -             \\ \hline

@@

We utilize the output from hidden layers 1, 3, 7, 12, 18, 23 from WavLM Large and Kmeans model with 
\(K=1000\).
We use 1024 as the embedding dimension.  The cross-attention module consists of 4 transformer encoders, each with 16 attention heads and an MLP with a hidden dimension of 1024. Layer normalization is applied after the cross-attention module.
The LM of small version TSELM-S uses 
embedding dimension \(d\) = 256, absolute sinusoidal positional embedding, conformer encoders as the backbone of LM. The conformer encoder consists of 6 layers with a kernel size of 31, each with 4 attention heads, and an MLP with a hidden dimension of 2048. The medium version TSELM-M uses \(d\) = 512 with 8 layers and 8 heads and the large version TSELM-L uses 
\(d\) = 768 with 12 layers and 16 heads. We use AdamW as 
our optimizer for all the experiments. The learning rate 
is \(5 \times 10^{-4}\) for TSELM-S and \(5 \times 10^{-5}\) for TSELM-M and TSELM-L. We train our models using 8 GPUs with 32 GB of RAM, each with batch size 16 for a total of 40,000 steps.

@@

\subsection{Baseline models}
Baseline models are presented in Table \ref{main_exp}. 
TSELM is compared with Spex+ \cite{spex_plus}, a discriminative separation model trained on Libri2Mix \cite{librimix}. 
We also compare our model with SkiM-UniCATS(vq-wav2vec) from \cite{gen_tse}, which uses discrete tokens from vq-wav2vec to conduct TSE. 
Mixture refers to the unprocessed mixed speech. 
Target-Discrete refers to the discretized target speech using our encoder and decoder. 
It serves as the upper bond for our model performance.  Besides TSELM, we conduct 3 main experiments, named
 Continuous-WavLM-L6, TSELM-L-Hybrid and TSELM-S-NoCat using the same training data. 
For Continuous-WavLM-L6, we directly
pass the embeddings from the 6th hidden layer output of WavLM Large to the cross-attention without discretization. 
The concatenation strategy is still applied to the mixed speech. Mean Square Error loss is applied between the output embeddings and the clean 
embeddings. HiFi-GAN in \cite{knn_vc} is used for audio reconstruction. For 
TSELM-L-Hybrid, inspired by MaskSR \cite{mask_sr}, we discretize the reference 
speech while utilizing the continuous embeddings from the mixed speech. In TSELM-S-NoCat, we utilize the TSELM-S model architecture but abandon the mixture concatenation strategy in the encoding step. The code and pretrained models are available at \footnote{\href{https://github.com/Beilong-Tang/TSELM}{https://github.com/Beilong-Tang/TSELM}}.

\section{Results and Discussions}
Table \ref{main_exp} presents the performance of different systems evaluated on the Libri2Mix testset and WSJ0-2mix testset. 
DNSMOS is computed over the 
output since this metric is reference-free. dWER is calculated with respect to the clean speech. 
For continuous methods, speaker similarity is directly computed against the clean speech. 
However, since discretization inherently results in some loss of speaker information—Target-Discrete shows a speaker similarity score of 0.654 on Libri2Mix —we follow \cite{dasb} to assess the speaker similarity of the output from the discrete methods against the target audio produced by discretizing the clean speech. 
The observed speaker information loss is likely due to the tokenization process, which inherently reduces speaker fidelity. Future work should aim to enhance tokenization methods for SSL models to mitigate this loss. Since the primary goal of this research is to explore target speaker extraction using discretized information rather than to develop improved tokenization methods, we argue that comparing the speaker similarity output with discretized clean speech is a reasonable approach, as it represents an upper bound of performance.

We observe that TSELM-L outperforms Spex+ in terms of DNSMOS scores, indicating better speech quality, but performs slightly worse in dWER, suggesting lower speech intelligibility. One potential reason for this could be the discretization process applied to the mixed speech. Our Kmeans algorithm is trained on clean speech rather than mixed speech, which is advantageous for speech enhancement as it likely aids in denoising. However, when applied to mixed speech, this discretization might lead to the model focusing on the wrong speaker, as it is most likely to retain only the dominant speaker information, causing a reduction in intelligibility.
This hypothesis is supported by the results of TSELM-L-Hybrid, where continuous embeddings from the mixture speech are used without discretization, achieving dWER scores better than Spex+ on WSJ0-2mix. 
% Another contributing factor could be the limitations of our current language model (LM). The encoder-only LM achieves around 55\% accuracy for token prediction on Libri2Mix testset, and we believe using more advanced models, such as auto-regressive or masking-based LMs, could lead to improved performance in future iterations.

@@

In Table \ref{hubert_wavlm_6}, we compare the performance of HuBERT and WavLM as SSL models and examine the effects of using either one or multiple layers for discretization. Our findings indicate that using HuBERT as the SSL model results in slightly better DNSMOS scores but much worse dWER compared to our WavLM baseline. The improved DNSMOS scores likely stem from the vocoder performance, yet they do not adequately reflect speech intelligibility, which is crucial for speech separation tasks.
The poorer dWER scores observed with this HuBERT model may be attributed to its training on clean speech, which might not equip it to capture the complexity and richness of mixed speech. Moreover, our results from WavLM-L6 suggest that when performing speech separation, discretizing across multiple layers provides better results than relying on a single layer.  This might be because that using multiple layer outputs can better tolerate errors compared to using just one layer output.

Finally, we observe a performance gap between discrete methods and continuous methods, as demonstrated by Continuous-WavLM-L6.
Continuous-WavLM-L6 has the best performance in terms of DNSMOS and dWER among all the experiments while using only the 6th layer output of WavLM Large. 
The gap in performance may be attributed to the information loss inherent in the discretization process. 
We hope our future research will bridge this gap.

\section{Conclusion}
In this work, we introduce a novel way using discrete tokens and language models for target speaker extraction.
Our method leverages multiple hidden layers of WavLM and Kmeans tokenizers for encoding, employs cross-attention and a language model for separation, and utilizes a scalable HiFi-GAN for audio reconstruction.
Experiments have 
shown that our model can achieve excellent performance in terms of speech quality, and comparable performance in terms of speech intelligibility. However, we observe a gap between discrete methods and continuous methods especially in speech intelligibility and speaker similarity. Our future work will focus on shrinking 
this gap.
